---
title: "Introduction to R and data visualisation"
subtitle: "The grammar of graphics · Day 1, Session 3"
author: "Gurpreet"
date: 2026-09-09
format:
  revealjs:
    theme: [default, ../custom.scss]
    slide-number: c/t
    footer: "MedEd Conclave 2026 · R & Reproducible Research"
    transition: fade
    center: false
execute:
  echo: true
  warning: false
---

```{r}
#| include: false
library(tidyverse)
theme_set(theme_minimal(base_size = 16))
students <- read_csv("data/meded_students.csv", show_col_types = FALSE)
```

## Look before you leap {.section-divider}

```{r}
#| output-location: fragment
glimpse(students)
```

::: {.notes}
⏱️ 3 min. Point out posttest is dbl with NAs. Foreshadow the warning that
appears on the next histogram.
:::

## Every plot is three things

::: {.big-number}
data + mapping + geometry
:::

```r
ggplot(data = students, aes(x = pretest, y = posttest)) +
  geom_point()
```

. . .

Layers join with `+`. Data flows with `|>`. Everyone confuses them for a week.

## One variable

```{r}
#| output-location: column-fragment
#| fig-width: 5
#| fig-height: 4
#| warning: true
ggplot(students, aes(posttest)) +
  geom_histogram(binwidth = 5,
                 fill = "#0f5257",
                 colour = "white")
```

::: {.notes}
🎯 STOP on the warning: "Removed 11 rows". Ask the room what it means before
telling them. This is the most important 60 seconds of the session.
:::

## Two variables

```{r}
#| output-location: column-fragment
#| fig-width: 5
#| fig-height: 4
ggplot(students,
       aes(teaching_arm, posttest,
           fill = teaching_arm)) +
  geom_boxplot(alpha = .7,
               outlier.shape = NA) +
  geom_jitter(width = .15, alpha = .3) +
  theme(legend.position = "none")
```

Box for the summary, points for the truth.

## Inside `aes()` or outside?

| | Means |
|---|---|
| `aes(colour = arm)` | Colour is **mapped** to a variable |
| `geom_point(colour = "navy")` | Colour is **fixed** decoration |

. . .

A legend with one meaningless entry = you put a constant inside `aes()`.

## Small multiples

```{r}
#| fig-height: 3.2
#| fig-width: 10
ggplot(students, aes(gain, fill = teaching_arm)) +
  geom_density(alpha = .6) +
  facet_wrap(~ college) +
  theme(legend.position = "bottom")
```

One line to check whether your finding holds across sites.

## Publication quality

```r
labs(title = "Knowledge gain by teaching method",
     subtitle = "Three colleges, 2026 (n = 229 complete)",
     y = "Gain (post − pre)", x = NULL,
     caption = "Synthetic data.")

ggsave("output/fig1.png", width = 7, height = 4.5, dpi = 300)
```

. . .

**Never screenshot a plot.**

## Hands-on — 60 min {background-color="#0f5257"}

Exercises 1–3 on the website. 4 if you fly.

Green sticky when Exercise 1 renders.

::: {.notes}
📋 Facilitators split the room in four. Common failures: `+` vs `|>`,
capitalisation, project not open.
:::
