---
title: "Working with data"
subtitle: "dplyr and exploratory data analysis · Day 1, Session 5"
author: "Gurpreet"
date: 2026-09-09
format:
  revealjs:
    theme: [default, ../custom.scss]
    slide-number: c/t
    footer: "MedEd Conclave 2026 · R & Reproducible Research"
    transition: fade
    center: false
execute:
  echo: true
  warning: false
---

```{r}
#| include: false
library(tidyverse)
students <- read_csv("data/meded_students.csv", show_col_types = FALSE)
osce <- read_csv("data/osce_stations.csv", show_col_types = FALSE)
```

## Six verbs {.section-divider}

| Verb | Acts on |
|---|---|
| `filter()` | rows |
| `select()` | columns |
| `arrange()` | order |
| `mutate()` | new columns |
| `group_by()` | strata |
| `summarise()` | rows → one |

Each takes a data frame, returns a data frame. That is why they chain.

## filter + select

```{r}
#| output-location: fragment
students |>
  filter(teaching_arm == "Flipped classroom", attendance_pct > 90) |>
  select(student_id, college, gain) |>
  head(3)
```

`==` not `=`. Comma means AND.

## The workhorse

```{r}
#| output-location: fragment
students |>
  group_by(teaching_arm) |>
  summarise(
    n = n(),
    n_gain = sum(!is.na(gain)),
    mean_gain = round(mean(gain, na.rm = TRUE), 2),
    .groups = "drop"
  )
```

::: {.notes}
🎯 Stop on na.rm. It is an exclusion decision, not a formality. Always report
n beside the mean so the exclusion is visible.
:::

## `na.rm = TRUE` is a decision

::: {.big-number}
11 students
:::

Without it: `NA`. With it: you have silently excluded eleven people.

**Both defensible. Neither automatic.**

## Joins — check the rows

```{r}
#| output-location: fragment
osce_summary <- osce |>
  group_by(student_id) |>
  summarise(osce_total = sum(score), .groups = "drop")

combined <- left_join(students, osce_summary, by = "student_id")
c(nrow(students), nrow(combined))
```

Row count grew? Your key is not unique. Silent, plausible, wrong.

## EDA — four questions, in order

::: {.incremental}
1. What am I holding? `glimpse()`, `skim()`
2. How is each variable distributed?
3. **Are the groups comparable to begin with?**
4. Only now: the comparison
:::

::: {.notes}
Question 3 is the one people skip. If the arms differ at baseline, no
downstream cleverness fixes it.
:::

## Don't go hunting

Twenty tests at α = 0.05 → one "significant" by construction.

. . .

Reproducible tooling makes p-hacking *easier*. That is exactly why the
discipline matters more, not less.

## Hands-on — 45 min {background-color="#0f5257"}

Exercises 1–3. Exercise 4 cleans the messy survey.

::: {.notes}
📋 Watch for: `=` vs `==`, forgotten `.groups`, and joins that duplicate.
:::
