---
title: "2 · Getting started"
subtitle: "Concepts, thumb rules, and data science & AI basics"
---

::: {.session-meta}
::: {.meta-item}
[Time]{.meta-label}[10:15 – 11:00]{.meta-value}
:::
::: {.meta-item}
[Faculty]{.meta-label}[Arun Mitra]{.meta-value}
:::
::: {.meta-item}
[Format]{.meta-label}[Demo + lecture]{.meta-value}
:::
::: {.meta-item}
[Slides]{.meta-label}[[Open deck ↗](../slides/day1/s2-getting-started.html){target="_blank"}]{.meta-value}
:::
:::

## Learning objectives

1. Navigate the four RStudio panes and know what each is for.
2. Structure a project folder so that it survives you.
3. Explain what a data science workflow does, end to end.
4. Place "AI" correctly relative to statistics, machine learning and
   generative models — and say what each is *for*.

::: {.callout-important}
## Opens with a 10-minute project-structure demo
Day 1 goes to `ggplot2` at 11:15. Before that, everyone needs a working
project and a data frame on screen. This session starts there. If you did the
[prelude](../prelude/install.qmd#projects), this is revision — help your
neighbour.
:::

## The RStudio window

| Pane | Where | Use it for |
|---|---|---|
| **Source** | top left | Your script. The thing you save. |
| **Console** | bottom left | Throwaway experiments. Nothing here is saved. |
| **Environment** | top right | What objects currently exist. Your reality check. |
| **Files / Plots / Help** | bottom right | Browsing, figures, `?function` |

The rule: **anything you would be sad to lose goes in Source.** The console
is scratch paper.

<kbd>Ctrl</kbd>/<kbd>Cmd</kbd> + <kbd>Enter</kbd> sends the current line from
Source to Console. That is the loop you will spend two days in.

## Project structure that survives you

```
meded-conclave-2026/
├── meded-conclave-2026.Rproj
├── data/                  # raw data. READ-ONLY. Never edit these files.
│   ├── meded_students.csv
│   └── osce_stations.csv
├── R/                     # scripts, numbered in run order
│   ├── 00_generate_datasets.R
│   ├── 01_clean.R
│   └── 02_analyse.R
├── output/                # everything R writes: figures, tables, reports
└── README.md              # what this is, who made it, how to run it
```

Three thumb rules:

1. **Raw data is read-only.** If you need to fix a typo in the data, fix it
   in a *script*, so the fix is visible and repeatable.
2. **Number your scripts** in the order they must run. `01_`, `02_`. Future
   you will not remember the order.
3. **Everything in `output/` must be regenerable.** If you deleted the folder
   and ran your scripts, it should come back identically. If it would not,
   something important is only in your head.

::: {.callout-tip}
## `here::here()`
Once your project has subfolders, `here::here("data", "meded_students.csv")`
builds the path correctly on any machine and from any subfolder. Worth
adopting once you are past the first day.
:::

## The data science workflow

```
Import → Tidy → [ Transform ⇄ Visualise ⇄ Model ] → Communicate
                          ↑____________________|
                        (the loop where the work is)
```

After Wickham & Grolemund. Mapped onto these two days:

| Stage | Package | When |
|---|---|---|
| Import | `readr` | Today, 11:15 |
| Tidy | `tidyr`, `janitor` | Today, 15:30 |
| Transform | `dplyr` | Today, 15:30 |
| Visualise | `ggplot2` | Today, 11:15 |
| Model | `stats`, `gtsummary` | Tomorrow, 11:15 |
| Communicate | `quarto` | Today 14:00, tomorrow 09:30 |

Note what dominates in practice: **import and tidy take most of the time**,
usually 60–80% of it. Nobody's methods section says so.

## Data science, machine learning, AI: what is what

These words are used interchangeably in seminars and they should not be.

| | Question it answers | Example in med ed |
|---|---|---|
| **Statistics** | Is this difference bigger than chance? | Does the flipped arm score higher on the post-test? |
| **Machine learning** | Can I predict `y` from `x` in new cases? | Predict which students will fail the OSCE |
| **Generative AI** | Can I produce plausible new text/code/images? | Draft the code for a figure; summarise free-text feedback |

Three distinctions worth holding on to:

- **Explanation vs prediction.** A model that predicts well may explain
  nothing, and vice versa. Most medical education research wants explanation
  and reaches for prediction language anyway.
- **Fitting vs generalising.** Any model can be made to fit the data you have.
  The question is always what it does on data you have not seen.
- **Fluency vs truth.** A generative model optimises for *plausible*, not
  *correct*. That distinction is the entire content of Day 2 afternoon.

::: {.callout-warning}
## Where AI fits in this workshop
As an accelerator on top of a workflow you understand — not a substitute for
understanding it. You cannot check code you could not have written. We return
to this properly in [Session 8](../day2/s8-genai.qmd).
:::

## Thumb rules to write on your hand

1. **Restart R often.** <kbd>Ctrl</kbd>/<kbd>Cmd</kbd>+<kbd>Shift</kbd>+<kbd>F10</kbd>.
   If your code only works in a warm session, it does not work.
2. **Look at the data.** `glimpse()`, `str()`, `View()`. Before every
   analysis, and after every step that changes the data.
3. **Small steps, checked.** Write one pipe step, run it, look at it, then
   add the next. Not five steps then debug.
4. **Read the error's first line.** R errors are verbose but the first line is
   usually the whole story.
5. **Name things for humans.** `posttest_by_arm`, not `df2`.
6. **If you did it twice, script it. Three times, write a function.**

## Quick check

::: {.callout-note title="Check your setup: 5 minutes"}
In your project, run:

```{r}
#| eval: false
getwd()                                       # should end in meded-conclave-2026
list.files("data")                            # should list 4 csv files
library(tidyverse)
students <- read_csv("data/meded_students.csv")
glimpse(students)                             # 240 rows, 17 columns
```

Red sticky note if any line fails. This is the last moment before we start
plotting.
:::

::: {.callout-tip title="If `read_csv()` cannot find the file" collapse="true"}
```{r}
#| eval: false
getwd()          # Are you where you think you are?
list.files()     # Is there a data/ folder here at all?
```

Almost always one of: (a) you opened the script but not the *project*, (b)
the files are in `Downloads/`, not `data/`, or (c) the file is called
`meded_students.csv.txt` because Windows hid the extension.
:::

## Key takeaways

::: {.callout-note appearance="minimal"}
- Source pane for anything you want to keep; console for scratch.
- Raw data read-only, scripts numbered, `output/` regenerable.
- Import and tidy are most of the work, whatever the methods section says.
- Statistics explains, ML predicts, generative AI produces plausible text.
  Choosing the wrong one is a study design error, not a software error.
:::

## Next

→ [3 · Introduction to R and data visualisation](s3-ggplot2.qmd)
