---
title: "3 · Introduction to R and data visualisation"
subtitle: "The grammar of graphics with ggplot2"
---

::: {.session-meta}
::: {.meta-item}
[Teaching]{.meta-label}[11:15 – 12:00]{.meta-value}
:::
::: {.meta-item}
[Hands-on]{.meta-label}[12:00 – 13:00]{.meta-value}
:::
::: {.meta-item}
[Faculty]{.meta-label}[Gurpreet]{.meta-value}
:::
::: {.meta-item}
[Slides]{.meta-label}[[Open deck ↗](../slides/day1/s3-ggplot2.html){target="_blank"}]{.meta-value}
:::
:::

```{r}
#| include: false
library(tidyverse)
theme_set(theme_minimal(base_size = 12))
students <- read_csv("data/meded_students.csv", show_col_types = FALSE)
```

## Learning objectives

1. Read a dataset into R and inspect it before touching it.
2. Explain the three components every `ggplot2` plot needs: data, aesthetic
   mapping, geometry.
3. Build histograms, boxplots and scatterplots of medical education data.
4. Split a plot by a grouping variable with `facet_wrap()`.
5. Label a figure so that it stands alone in a manuscript.

## Start by looking

```{r}
library(tidyverse)
students <- read_csv("data/meded_students.csv")
```

```{r}
glimpse(students)
```

240 rows, 17 columns, one row per student. Note `posttest` is `dbl` with
missing values — we will meet them shortly.

## The grammar of graphics in one idea

Every plot is the same three things:

> **data** + **a mapping from variables to visual properties** + **a geometry that draws it**

In code:

```{r}
#| eval: false
ggplot(data = students, mapping = aes(x = pretest, y = posttest)) +
  geom_point()
```

- `ggplot()` opens a canvas and says which data.
- `aes()` maps *variables* to *visual channels* — x position, y position,
  colour, size, shape.
- `geom_*()` chooses how to draw it.
- Layers are joined with `+`, not `|>`. This is a historical quirk; everyone
  gets it wrong for the first week.

## One variable: what does the distribution look like?

```{r}
#| warning: true
ggplot(students, aes(x = posttest)) +
  geom_histogram(binwidth = 5, fill = "#0f5257", colour = "white")
```

::: {.callout-important}
## That warning is the point
`Removed 11 rows containing non-finite values`. Eleven students have no
post-test score. `ggplot2` told you, quietly, in a message most people
scroll past. Real analyses die here.

Always read the warning. Then decide, explicitly, what to do about the
missing eleven — and write that decision into your script.
:::

## Two variables, one categorical: does the arm matter?

```{r}
ggplot(students, aes(x = teaching_arm, y = posttest, fill = teaching_arm)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c("Flipped classroom" = "#0f5257",
                               "Traditional lecture" = "#b8860b")) +
  labs(x = NULL, y = "Post-test score (0–100)") +
  theme(legend.position = "none")
```

Boxplot for the summary, jittered points for the raw data. Always show the
points when *n* is small enough to allow it — a boxplot of 12 observations
hides more than it reveals.

## Two continuous variables

```{r}
ggplot(students, aes(x = pretest, y = posttest, colour = teaching_arm)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_colour_manual(values = c("Flipped classroom" = "#0f5257",
                                 "Traditional lecture" = "#b8860b")) +
  labs(x = "Pre-test score", y = "Post-test score", colour = "Teaching arm")
```

Adding `colour = teaching_arm` inside `aes()` splits *everything* downstream —
points and both fitted lines. That is the grammar doing the work.

::: {.callout-tip}
## `colour =` inside vs outside `aes()`
- Inside `aes()`: colour **means** something — it is mapped to a variable.
- Outside `aes()`: colour is just decoration — `geom_point(colour = "navy")`.

Putting a fixed colour inside `aes()` produces the classic confusing result of
a legend with one meaningless entry.
:::

## Small multiples

```{r}
#| fig-height: 3.2
ggplot(students, aes(x = gain, fill = teaching_arm)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ college, ncol = 3) +
  scale_fill_manual(values = c("Flipped classroom" = "#0f5257",
                               "Traditional lecture" = "#b8860b")) +
  labs(x = "Knowledge gain (post − pre)", y = "Density", fill = NULL) +
  theme(legend.position = "bottom")
```

`facet_wrap()` repeats the same plot across levels of a variable. It is the
single highest-value function in `ggplot2` for exploratory work: one line,
and you have checked whether your finding holds across sites.

## Making it publishable

```{r}
#| fig-height: 4
ggplot(students, aes(x = teaching_arm, y = gain, fill = teaching_arm)) +
  geom_boxplot(alpha = 0.75, width = 0.55) +
  scale_fill_manual(values = c("Flipped classroom" = "#0f5257",
                               "Traditional lecture" = "#b8860b")) +
  labs(
    title    = "Knowledge gain by teaching method",
    subtitle = "Three government medical colleges, 2026 (n = 229 with complete data)",
    x        = NULL,
    y        = "Gain in knowledge score (post − pre)",
    caption  = "Synthetic data for teaching purposes."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

Then save it — at a size and resolution a journal will accept:

```{r}
#| eval: false
ggsave("output/fig1_gain_by_arm.png",
       width = 7, height = 4.5, dpi = 300, bg = "white")
```

::: {.callout-note}
## Never screenshot a plot
`ggsave()` writes exactly what you specify, every time you run the script. A
screenshot is a manual step that has to be repeated by hand whenever the data
change — and it will be the wrong resolution.
:::

# Hands-on {#hands-on}

**12:00 – 13:00.** Work in your groups. Facilitators are circulating. Do
Exercises 1–3; Exercise 4 if you have time.

::: {.callout-note title="Exercise 1: first plot, guaranteed success"}
**Time:** 10 minutes

1. Load `students` if you have not already.
2. Make a histogram of `attendance_pct`.
3. Change the `binwidth` to 2, then to 10. Which tells you more?
4. Give it an axis label and a title with `labs()`.
:::

::: {.callout-tip title="Solution 1" collapse="true"}
```{r}
#| fig-height: 3
ggplot(students, aes(x = attendance_pct)) +
  geom_histogram(binwidth = 2, fill = "#0f5257", colour = "white") +
  labs(title = "Attendance across the module",
       x = "Attendance (%)", y = "Number of students")
```

`binwidth = 2` shows the shape and the left tail; `binwidth = 10` smooths the
tail away. There is no correct answer, only a decision you should be able to
defend — which is why the default bin count comes with a warning.
:::

::: {.callout-note title="Exercise 2: comparison"}
**Time:** 15 minutes

Does burnout differ by teaching arm?

1. Boxplot of `burnout_score` by `teaching_arm`.
2. Add the raw points with `geom_jitter()`.
3. Facet by `sex`.
4. What do you conclude — and what would you need to check before believing
   it?
:::

::: {.callout-tip title="Solution 2" collapse="true"}
```{r}
#| fig-height: 3.4
ggplot(students, aes(x = teaching_arm, y = burnout_score)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, fill = "#4a7c7e") +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1) +
  facet_wrap(~ sex) +
  labs(x = NULL, y = "Burnout score (0–100)")
```

The distributions overlap almost completely. Burnout was generated from
attendance, not from the teaching arm — so the honest conclusion is "no
visible difference". Before believing *any* difference you saw, you would
check: how many students per cell, how much missingness (7 here), and whether
the arms differ in attendance to begin with.

That last question is confounding, and no plot will answer it for you.
:::

::: {.callout-note title="Exercise 3: relationship"}
**Time:** 15 minutes

1. Scatterplot of `study_hours` (x) against `gain` (y).
2. Colour the points by `rural_background`.
3. Add a straight fitted line per group.
4. Facet by `batch`.
5. Label everything properly and save it to `output/` with `ggsave()`.
:::

::: {.callout-tip title="Solution 3" collapse="true"}
```{r}
#| fig-height: 3.2
ggplot(students, aes(x = study_hours, y = gain, colour = rural_background)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ batch) +
  scale_colour_manual(values = c("Rural" = "#0f5257", "Urban" = "#b8860b")) +
  labs(x = "Self-reported study hours per week",
       y = "Knowledge gain",
       colour = "Background")
```

```{r}
#| eval: false
ggsave("output/study_hours_gain.png", width = 8, height = 4, dpi = 300, bg = "white")
```

The lines are near-flat. `study_hours` was generated independently of `gain` —
so a flat line is the correct answer, and finding it is a useful experience.
Most exploratory plots show nothing. Publication bias is partly the habit of
only remembering the ones that did.
:::

::: {.callout-note title="Exercise 4 (stretch): the OSCE data"}
**Time:** 15 minutes

Use `osce_stations.csv` (1,200 rows, five per student).

1. Read it in and `glimpse()` it.
2. Boxplot of `score` by `station_name`. Rotate the x labels so they are
   readable — or use `coord_flip()`.
3. Colour by `domain`.
4. Which station is hardest? Does that match what you would expect?
:::

::: {.callout-tip title="Solution 4" collapse="true"}
```{r}
#| fig-height: 3.4
osce <- read_csv("data/osce_stations.csv", show_col_types = FALSE)

osce |>
  mutate(station_name = fct_reorder(station_name, score)) |>
  ggplot(aes(x = station_name, y = score, fill = domain)) +
  geom_boxplot(alpha = 0.8) +
  coord_flip() +
  labs(x = NULL, y = "Station score (out of 20)", fill = "Domain")
```

`fct_reorder()` sorts the stations by median score — always order a
categorical axis by something meaningful, never alphabetically.

The procedural station (IV access) is hardest, then data interpretation.
Whether that reflects student ability or examiner stringency is not
answerable from this plot, and that ambiguity is itself an important
finding about station-level assessment.
:::

## Key takeaways

::: {.callout-note appearance="minimal"}
- Three components: `ggplot(data, aes(...)) + geom_*()`. Everything else is
  refinement.
- Inside `aes()` means "mapped to a variable"; outside means "fixed".
- Layers join with `+`. Pipes join data operations with `|>`.
- `facet_wrap()` is the cheapest way to check whether a finding holds across
  subgroups.
- Read the warnings. `Removed 11 rows` is a finding, not noise.
- `ggsave()`, never a screenshot.
:::

## Next

→ [4 · Introduction to Quarto](s4-quarto.qmd) — put these plots inside a
document that regenerates itself.
