---
title: "5 · Working with data"
subtitle: "dplyr verbs and exploratory data analysis"
---

::: {.session-meta}
::: {.meta-item}
[Teaching]{.meta-label}[15:30 – 16:15]{.meta-value}
:::
::: {.meta-item}
[Hands-on]{.meta-label}[16:15 – 17:00]{.meta-value}
:::
::: {.meta-item}
[Faculty]{.meta-label}[Gurpreet]{.meta-value}
:::
::: {.meta-item}
[Slides]{.meta-label}[[Open deck ↗](../slides/day1/s5-dplyr-eda.html){target="_blank"}]{.meta-value}
:::
:::

```{r}
#| include: false
library(tidyverse)
theme_set(theme_minimal(base_size = 12))
students <- read_csv("data/meded_students.csv", show_col_types = FALSE)
osce     <- read_csv("data/osce_stations.csv", show_col_types = FALSE)
```

## Learning objectives

1. Use the six core `dplyr` verbs: `filter`, `select`, `arrange`, `mutate`,
   `group_by`, `summarise`.
2. Chain them with `|>` into a readable pipeline.
3. Handle missing values deliberately rather than by accident.
4. Join two tables on a shared key.
5. Run and interpret a basic exploratory analysis before modelling anything.

## Six verbs

Every one takes a data frame as its first argument and returns a data frame.
That is what makes them chainable.

| Verb | Acts on | Question |
|---|---|---|
| `filter()` | rows | Which observations do I want? |
| `select()` | columns | Which variables do I want? |
| `arrange()` | rows | In what order? |
| `mutate()` | columns | What new variable do I need? |
| `group_by()` | groups | Within which strata? |
| `summarise()` | rows → 1 | What summary do I want? |

### `filter()`: choose rows

```{r}
students |>
  filter(teaching_arm == "Flipped classroom", attendance_pct > 90) |>
  nrow()
```

`==` for equality (a single `=` is assignment and will error). Commas mean
AND; use `|` for OR. `%in%` for a set:

```{r}
students |>
  filter(batch %in% c("2023", "2024")) |>
  nrow()
```

### `select()`: choose columns

```{r}
students |>
  select(student_id, teaching_arm, pretest, posttest, gain) |>
  head(4)
```

Helpers worth knowing: `starts_with("q")`, `ends_with("_score")`,
`contains("test")`, `where(is.numeric)`.

### `mutate()`: create or change columns

```{r}
students |>
  mutate(
    gain_pct  = round(100 * gain / pretest, 1),
    high_risk = attendance_pct < 75
  ) |>
  select(student_id, pretest, gain, gain_pct, high_risk) |>
  head(4)
```

`if_else()` and `case_when()` for conditional recoding:

```{r}
students |>
  mutate(
    attendance_band = case_when(
      attendance_pct >= 90 ~ "High",
      attendance_pct >= 75 ~ "Adequate",
      .default = "Below requirement"
    )
  ) |>
  count(attendance_band)
```

### `group_by()` + `summarise()`: the workhorse

```{r}
students |>
  group_by(teaching_arm) |>
  summarise(
    n         = n(),
    n_gain    = sum(!is.na(gain)),
    mean_gain = mean(gain, na.rm = TRUE),
    sd_gain   = sd(gain, na.rm = TRUE),
    median_gain = median(gain, na.rm = TRUE)
  )
```

::: {.callout-important}
## `na.rm = TRUE` is a decision, not a formality
Without it, `mean(gain)` returns `NA`, R refusing to guess. With it, you
have silently excluded 11 students. **Both are defensible; neither is
automatic.** Always report `n` alongside the mean, as above, so the exclusion
is visible to your reader.
:::

Group by more than one variable:

```{r}
students |>
  group_by(college, teaching_arm) |>
  summarise(n = n(), mean_gain = round(mean(gain, na.rm = TRUE), 1),
            .groups = "drop")
```

`.groups = "drop"` ungroups the result. Without it, the grouping persists and
your next `mutate()` behaves in ways you did not intend: a bug that produces
plausible wrong numbers rather than an error.

## Missing data, deliberately

```{r}
students |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0)
```

Three columns have missing values, all traceable to the 11 students without a
post-test plus 7 without a burnout score. Now you can decide (and document)
what to do. What you must not do is discover it in a footnote after
submission.

## Joining tables

Student-level data in one file, station-level in another. Bring them
together on `student_id`:

```{r}
osce_summary <- osce |>
  group_by(student_id) |>
  summarise(osce_total = sum(score), osce_mean = mean(score), .groups = "drop")

combined <- students |>
  left_join(osce_summary, by = "student_id")

combined |>
  select(student_id, teaching_arm, posttest, osce_total) |>
  head(4)
```

`left_join()` keeps every row of the left table. Always check the row count
afterwards:

```{r}
nrow(students); nrow(combined)
```

If the second number is larger, your key is not unique and you have silently
duplicated rows. This is the most common and most damaging error in applied
data work.

## An honest exploratory analysis

EDA is not a step before the real analysis. It *is* most of the analysis.
Four questions, in order:

**1. What am I holding?**

```{r}
#| eval: false
glimpse(students)
skimr::skim(students)
```

**2. How is each variable distributed?**

```{r}
#| fig-height: 2.8
students |>
  select(pretest, posttest, attendance_pct, burnout_score) |>
  pivot_longer(everything()) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 25, fill = "#0f5257") +
  facet_wrap(~ name, scales = "free")
```

**3. Are the groups comparable to begin with?**

```{r}
students |>
  group_by(teaching_arm) |>
  summarise(
    n = n(),
    mean_pretest    = round(mean(pretest), 1),
    mean_attendance = round(mean(attendance_pct), 1),
    pct_female      = round(100 * mean(sex == "Female"), 1),
    .groups = "drop"
  )
```

If the arms differ at baseline, any difference in outcome is confounded, and
no amount of downstream cleverness fixes it. Check first.

**4. Only now, the comparison.**

```{r}
students |>
  group_by(teaching_arm) |>
  summarise(n = sum(!is.na(gain)),
            mean_gain = round(mean(gain, na.rm = TRUE), 2),
            se = round(sd(gain, na.rm = TRUE) / sqrt(n), 2),
            .groups = "drop")
```

```{r}
t.test(gain ~ teaching_arm, data = students)
```

::: {.callout-warning}
## Do not go hunting
Having found one comparison, it is tempting to try every other one. Run
twenty tests at α = 0.05 and one will be "significant" by construction. If
you did not pre-specify it, describe it as exploratory and say so in the
paper. Reproducible tooling makes this *easier* to do dishonestly, and
therefore more important to do honestly.
:::

# Hands-on {#hands-on}

**16:15 – 17:00.** Groups. Exercises 1–3 for everyone; 4 if you finish.

::: {.callout-note title="Exercise 1: filter, select, arrange"}
**Time:** 10 minutes

1. How many students in the 2024 batch attended more than 85%?
2. Make a table of the ten students with the highest `gain`, showing only
   `student_id`, `college`, `teaching_arm` and `gain`.
3. How many students are there per college? (`count()`)
:::

::: {.callout-tip title="Solution 1" collapse="true"}
```{r}
students |>
  filter(batch == "2024", attendance_pct > 85) |>
  nrow()

students |>
  select(student_id, college, teaching_arm, gain) |>
  arrange(desc(gain)) |>
  head(10)

students |> count(college)
```

`arrange(desc(gain))` puts `NA` last regardless of direction, R's sensible
default, and a good thing to know before you report a "top ten".
:::

::: {.callout-note title="Exercise 2: group and summarise"}
**Time:** 15 minutes

1. Mean and SD of `posttest` by `college`, with the *n* contributing to each
   mean.
2. Add `rural_background` as a second grouping variable. Are any cells small?
3. Which college × arm combination has the highest mean gain? Would you
   report that finding?
:::

::: {.callout-tip title="Solution 2" collapse="true"}
```{r}
students |>
  group_by(college) |>
  summarise(n = sum(!is.na(posttest)),
            mean = round(mean(posttest, na.rm = TRUE), 1),
            sd   = round(sd(posttest, na.rm = TRUE), 1),
            .groups = "drop")

students |>
  group_by(college, rural_background) |>
  summarise(n = n(), mean_gain = round(mean(gain, na.rm = TRUE), 1),
            .groups = "drop") |>
  arrange(n)
```

Some cells fall to around 25 students. A mean from 25 observations with an SD
near 8 has a standard error above 1.5: differences of a point or two between
such cells are noise.

On the third question: no. With six college × arm cells you are choosing the
maximum of six noisy estimates, which is guaranteed to overstate the effect.
Report the pre-specified overall comparison and describe the rest as
exploratory.
:::

::: {.callout-note title="Exercise 3: join and analyse"}
**Time:** 15 minutes

1. From `osce`, compute each student's mean score **per domain**.
2. Pivot it wider so there is one row per student and one column per domain
   (`pivot_wider()`).
3. Join to `students`.
4. Do flipped-classroom students score higher on the Communication domain?
   Summarise, then plot.
:::

::: {.callout-tip title="Solution 3" collapse="true"}
```{r}
domain_scores <- osce |>
  group_by(student_id, domain) |>
  summarise(mean_score = mean(score), .groups = "drop") |>
  pivot_wider(names_from = domain, values_from = mean_score) |>
  janitor::clean_names()

combined <- students |> left_join(domain_scores, by = "student_id")

combined |>
  group_by(teaching_arm) |>
  summarise(n = n(),
            communication = round(mean(communication), 2),
            cognitive     = round(mean(cognitive), 2),
            .groups = "drop")
```

```{r}
#| fig-height: 3
combined |>
  ggplot(aes(x = teaching_arm, y = communication, fill = teaching_arm)) +
  geom_boxplot(alpha = 0.75) +
  labs(x = NULL, y = "Mean communication-domain score (0–20)") +
  theme(legend.position = "none")
```

There is a small advantage in the communication domain, around 0.9 marks,
which is how the data were generated. It is real in this dataset and it is
also small relative to the spread. Reporting the effect size next to the
p-value, rather than instead of it, is the honest presentation.
:::

::: {.callout-note title="Exercise 4 (stretch): clean the messy survey"}
**Time:** 15 minutes

`data/faculty_survey_raw.csv` is a raw form export.

1. Read it and look at `names()`. Fix them with `janitor::clean_names()`.
2. Remove exact duplicate rows (`distinct()`). How many were there?
3. `department` has inconsistent spellings and stray whitespace. Standardise
   it with `str_trim()`, `str_to_title()` and a `case_when()` for
   Paediatrics/Pediatrics.
4. `ever_used_r` is coded `Yes/yes/No/no/N/""`. Recode to a clean logical.
5. Count respondents by department.
:::

::: {.callout-tip title="Solution 4" collapse="true"}
```{r}
faculty <- read_csv("data/faculty_survey_raw.csv", show_col_types = FALSE) |>
  janitor::clean_names()

names(faculty)

nrow(faculty)
faculty <- distinct(faculty)
nrow(faculty)     # 12 duplicates removed

faculty <- faculty |>
  mutate(
    department = department |> str_trim() |> str_to_title(),
    department = case_when(
      department %in% c("Pediatrics", "Paediatrics") ~ "Paediatrics",
      .default = department
    ),
    ever_used_r = case_when(
      str_to_lower(str_trim(ever_used_r)) %in% c("yes", "y") ~ TRUE,
      str_to_lower(str_trim(ever_used_r)) %in% c("no", "n")  ~ FALSE,
      .default = NA
    )
  )

faculty |> count(department, sort = TRUE)
```

Note the order: clean names first, deduplicate second, recode third. Recoding
before deduplicating means doing the work twice, and if your recoding is
imperfect, "duplicates" that differ only in whitespace will survive.
:::

## Key takeaways

::: {.callout-note appearance="minimal"}
- Six verbs, one pipe, arbitrarily complex analysis.
- `na.rm = TRUE` is an exclusion decision. Report `n` beside every mean.
- Check row counts after every join.
- `.groups = "drop"` after `summarise()`, unless you meant to stay grouped.
- Check baseline comparability *before* comparing outcomes.
- Clean names → deduplicate → recode. In that order.
:::

## End of Day 1

You have imported, visualised, wrangled and reported. Tomorrow: making it
publishable, and making it fast.

→ [Day 2](../day2/index.qmd)
