---
title: "Check your setup"
subtitle: "Packages and verification"
---

## Install the workshop packages

Open your `meded-conclave-2026` project in RStudio. Paste this into the
**Console** and press <kbd>Enter</kbd>. It will take 10–20 minutes on a first
run. Start it and make tea.

```{r}
#| eval: false
pkgs <- c(
  
  # Core
  "tidyverse",     # dplyr, ggplot2, tidyr, readr, stringr, forcats
  "here",          # sane file paths
  "janitor",       # clean_names(), tabyl()
  
  # Tables
  "gtsummary",     # publication-ready tables
  "gt",            # the table engine underneath
  "flextable",     # tables for Word output
  
  # Reporting
  "quarto",        # render .qmd from R
  "knitr",
  "rmarkdown",
  "rstudioapi",    # lets the setup check read your RStudio version

  # Data and extras
  "skimr",         # fast data overviews
  "scales",        # axis formatting
  "patchwork",     # combine ggplots
  "MASS"           # classic datasets from Venables & Ripley
)

install.packages(setdiff(pkgs, rownames(installed.packages())))
```

::: {.callout-warning}
## One thing to know about `MASS`
`MASS` has its own `select()`, so loading it *after* dplyr silently replaces
the dplyr one and `select(students, batch)` starts failing with a confusing
error. Load `MASS` first, or leave it unloaded and reach for a dataset with
`MASS::Boston`. If `select()` ever stops behaving, run `conflicts()`.
:::

::: {.callout-warning}
## If you are asked "Do you want to install from sources...?"
Answer **No** (type `n`, press Enter). Binary packages install in seconds;
source packages compile for minutes and sometimes fail. You do not need the
newest possible version for this workshop.
:::

## Run the verification script

Download [`check_setup.R`](../setup/check_setup.R), put it in your project's
`R/` folder, then run:

```{r}
#| eval: false
source("R/check_setup.R")
```

The script checks eight things: your R version, the native pipe, RStudio,
Quarto, all sixteen required packages, whether the core stack loads, whether
your project folder is writable, and whether Quarto and R can complete an
end-to-end render together.

## What a healthy result looks like

```
── MedEd Conclave 2026 · setup check ───────────────────────────
✔ R version 4.6.1 (2026-06-24) (need >= 4.1.0)
✔ Native pipe |> works
✔ RStudio 2026.08.0
✔ Quarto 1.10.18
✔ All 16 required packages installed
✔ Core packages load cleanly
✔ Project directory writable (meded-conclave-2026)
✔ Test render (HTML) succeeded
────────────────────────────────────────────────────────────────
You are ready for 9 September. See you there.
```

Any line starting with `✖` needs attention before the workshop. Send the
**whole** output to the SET Team if you cannot resolve it. The individual
lines tell us exactly what is wrong, which turns five emails into one.

## Test that Quarto renders

The single best proof that everything works end to end:

1. **File → New File → Quarto Document**
2. Title it `test`, choose **HTML**, click **Create**
3. Click **Render** (or <kbd>Ctrl</kbd>/<kbd>Cmd</kbd> + <kbd>Shift</kbd> + <kbd>K</kbd>)

A browser tab should open with a formatted document containing a plot. If it
does, your entire toolchain works and nothing on Day 1 can be blocked by
installation.

::: {.callout-tip}
## Save the evidence
Take a screenshot of a successful render and bring it. If something breaks on
the day, knowing it worked last week narrows the problem down enormously.
:::

## Common results and what they mean

| Line | Means | Do this |
|---|---|---|
| `✖ R version ... too old` | R below 4.1, so the native pipe will not work | Install R 4.6.x from CRAN |
| `✖ Quarto not found` | Quarto is not on your `PATH` | Reinstall Quarto, **restart the computer**, then `install.packages("quarto")` |
| `✖ Quarto X is installed, but the quarto R package is not` | Quarto itself is fine; R just cannot talk to it | `install.packages("quarto")` |
| `✖ N package(s) missing` | An install silently failed | Re-run the install line the script prints |
| `✖ Cannot write to ...` | Project is in a restricted or cloud-synced folder | Move it to Desktop or Documents |
| `✖ Test render failed` | R and Quarto are both present but not co-operating | See [Troubleshooting](../resources/troubleshooting.qmd) |
| `! Cannot read the RStudio version` | The `rstudioapi` package is missing | `install.packages("rstudioapi")` |
| `! Not running inside RStudio` | Not an error | Ignore it if you use another editor |

## Send us your result

Do this even if every line passed. It is how we know who is ready and who
needs help before 9 September.

1. Select the **whole** setup-check output in the Console and copy it, or take
   a screenshot of it.
2. Go to the [Moodle course](https://setfacilitysaral2.org/course/view.php?id=394),
   open **Get your laptop ready → Show us your setup check passed**.
3. Paste the output in the text box, or attach the screenshot. Then submit.

Something failed and you cannot fix it? Submit it anyway, exactly as it is,
and add one line saying which operating system you are on. A failed check we
can see on 25 August is a five-minute fix. The same failure discovered at
09:15 on 9 September costs you the first session.

## Next

→ [R primer](r-primer.qmd): The Basics of R so that you are ready for Day 1!!
