---
title: "Check your setup"
subtitle: "Packages and verification"
---

## Install the workshop packages

Open your `meded-conclave-2026` project in RStudio. Paste this into the
**Console** and press <kbd>Enter</kbd>. It will take 10–20 minutes on a first
run — start it and make tea.

```{r}
#| eval: false
pkgs <- c(
  # Core
  "tidyverse",     # dplyr, ggplot2, tidyr, readr, stringr, forcats, purrr
  "here",          # sane file paths
  "janitor",       # clean_names(), tabyl()
  # Tables
  "gtsummary",     # publication-ready tables
  "gt",            # the table engine underneath
  "flextable",     # tables for Word output
  # Reporting
  "quarto",        # render .qmd from R
  "knitr",
  "rmarkdown",
  # Data and extras
  "skimr",         # fast data overviews
  "scales",        # axis formatting
  "patchwork",     # combine ggplots
  "palmerpenguins" # a friendly practice dataset
)

install.packages(setdiff(pkgs, rownames(installed.packages())))
```

::: {.callout-warning}
## If you are asked "Do you want to install from sources...?"
Answer **No** (type `n`, press Enter). Binary packages install in seconds;
source packages compile for minutes and sometimes fail. You do not need the
newest possible version for this workshop.
:::

## Run the verification script

Download [`check_setup.R`](../setup/check_setup.R), put it in your project's
`R/` folder, then run:

```{r}
#| eval: false
source("R/check_setup.R")
```

The script checks eight things: your R version, the native pipe, RStudio,
Quarto, all sixteen required packages, whether the core stack loads, whether
your project folder is writable, and whether Quarto and R can complete an
end-to-end render together.

## What a healthy result looks like

```
── MedEd Conclave 2026 · setup check ───────────────────────────
✔ R version 4.4.1 (2024-06-14) (need >= 4.1.0)
✔ Native pipe |> works
✔ RStudio 2024.09.0+375
✔ Quarto 1.5.57
✔ All 16 required packages installed
✔ Core packages load cleanly
✔ Project directory writable (meded-conclave-2026)
✔ Test render (HTML) succeeded
────────────────────────────────────────────────────────────────
You are ready for 9 September. See you there.
```

Any line starting with `✖` needs attention before the workshop. Send the
**whole** output to the SET Team if you cannot resolve it — the individual
lines tell us exactly what is wrong, which turns five emails into one.

## Test that Quarto renders

The single best proof that everything works end to end:

1. **File → New File → Quarto Document**
2. Title it `test`, choose **HTML**, click **Create**
3. Click **Render** (or <kbd>Ctrl</kbd>/<kbd>Cmd</kbd> + <kbd>Shift</kbd> + <kbd>K</kbd>)

A browser tab should open with a formatted document containing a plot. If it
does, your entire toolchain works and nothing on Day 1 can be blocked by
installation.

::: {.callout-tip}
## Save the evidence
Take a screenshot of a successful render and bring it. If something breaks on
the day, knowing it worked last week narrows the problem down enormously.
:::

## Common results and what they mean

| Line | Means | Do this |
|---|---|---|
| `✖ R version ... too old` | R below 4.1, so the native pipe will not work | Install R 4.4.x from CRAN |
| `✖ Quarto not found` | Quarto is not on your `PATH` | Reinstall Quarto, **restart the computer**, then `install.packages("quarto")` |
| `✖ N package(s) missing` | An install silently failed | Re-run the install line the script prints |
| `✖ Cannot write to ...` | Project is in a restricted or cloud-synced folder | Move it to Desktop or Documents |
| `✖ Test render failed` | R and Quarto are both present but not co-operating | See [Troubleshooting](../resources/troubleshooting.qmd) |
| `! Not running inside RStudio` | Not an error | Ignore it if you use another editor |

## Next

→ [R primer](r-primer.qmd) — the 45 minutes of R that Day 1 assumes you have
already seen.
