---
title: "Datasets"
subtitle: "Three synthetic medical education datasets"
---

::: {.callout-important}
## These data are synthetic
Every value is generated by [`R/00_generate_datasets.R`](../R/00_generate_datasets.R)
with `set.seed(20260909)`. No real student, faculty member or institution is
represented. The structure and relationships are plausible for an Indian
medical college cohort — which is what makes them useful for teaching, and
useless for publication.
:::

## Download

Save all four files into the `data/` folder of your RStudio project.

| File | Rows | What it is |
|---|---|---|
| [`meded_students.csv`](../data/meded_students.csv) | 240 | One row per student: demographics, teaching arm, pre/post-test, burnout |
| [`osce_stations.csv`](../data/osce_stations.csv) | 1,200 | One row per student × OSCE station: scores, examiner, domain |
| [`faculty_survey_raw.csv`](../data/faculty_survey_raw.csv) | 132 | A deliberately messy Google Form export, for cleaning practice |
| [`codebook.csv`](../data/codebook.csv) | — | Variable dictionary for all of the above |

Or, if you would rather generate them yourself:

```{r}
#| eval: false
source("R/00_generate_datasets.R")
```

## The teaching scenario

Three government medical colleges ran a module on evidence-based practice.
Students were allocated to one of two teaching arms — **traditional lecture**
or **flipped classroom** — and assessed with a knowledge test before and
after, plus a five-station OSCE. A parallel survey asked faculty about their
own research and data practices.

That single scenario carries the whole workshop: `ggplot2` on Day 1 morning,
`dplyr` on Day 1 afternoon, `gtsummary` on Day 2 morning, and the group
activity on Day 2 afternoon.

## `meded_students.csv`

```{r}
#| eval: false
library(tidyverse)
students <- read_csv("data/meded_students.csv")
glimpse(students)
```

Key variables:

| Variable | Type | Notes |
|---|---|---|
| `student_id` | chr | Links to `osce_stations.csv` |
| `teaching_arm` | chr | `Traditional lecture` / `Flipped classroom` — the exposure |
| `pretest`, `posttest` | dbl | Knowledge, 0–100. **`posttest` has 11 missing values** |
| `gain` | dbl | `posttest − pretest`, missing wherever `posttest` is |
| `attendance_pct` | dbl | 45–100 |
| `burnout_score` | dbl | 0–100, higher is worse. **7 missing** |
| `burnout_band` | chr | Low / Moderate / High |
| `satisfaction` | dbl | 1–5 Likert |
| `passed` | chr | Yes if `posttest ≥ 50` |

::: {.callout-note}
## The missingness is deliberate
Eleven students have no post-test. You will meet them the first time you run
`mean(students$posttest)` and get `NA`. That is not a bug in the data; it is
the most common surprise in real analysis, and it is better to meet it here.
:::

## `osce_stations.csv`

Long format — five rows per student, one per station. This is the file we use
to teach `group_by()`, `summarise()` and joins.

| Station | Name | Domain |
|---|---|---|
| ST1 | History taking | Communication |
| ST2 | Clinical examination | Clinical skill |
| ST3 | Procedure: IV access | Procedural skill |
| ST4 | Counselling | Communication |
| ST5 | Data interpretation | Cognitive |

Scores are out of 20. Procedural and cognitive stations are marked harder —
a real pattern worth discovering rather than being told.

## `faculty_survey_raw.csv`

This one is ugly on purpose. It contains:

- Column names with spaces, punctuation and question text
- `Department` recorded as `Paediatrics`, `Pediatrics`, `physiology`, ` Anatomy`
- Dates in four different formats
- Missing values coded as `""`, `NA`, `-` and `unknown`
- Twelve duplicated rows
- Free-text comments with leading and trailing whitespace

If your cleaned version has 120 rows and tidy `snake_case` names, you have
won.

## Loading everything at once

```{r}
#| eval: false
library(tidyverse)

students <- read_csv("data/meded_students.csv")
osce     <- read_csv("data/osce_stations.csv")
faculty  <- read_csv("data/faculty_survey_raw.csv")

# Sanity check
dim(students)   # 240 x 17
dim(osce)       # 1200 x 8
dim(faculty)    # 132 x 11
```

If those dimensions do not match, you have an incomplete download. Fetch the
files again rather than debugging your code.

## Next

→ [Pre-reading](pre-reading.qmd)
