---
title: "R Primer"
subtitle: "The reference card: types, paths and the errors you will meet"
---

The workshop assumes you are already sensitized to R and RStudio. The teaching
happens in the five interactive primers, which run R **in your browser** with
nothing installed:

| | Primer | It teaches | Quiz |
|---|---|---|---|
| 1 | [Objects, functions and vectors](primer-01-objects.qmd) | `<-`, calling a function, `c()`, one column of values | [Primer 1](https://setfacilitysaral2.org/mod/quiz/view.php?id=3457) |
| 2 | [Tibbles](primer-02-tibbles.qmd) | The data rectangle: rows are observations, columns are variables | [Primer 2](https://setfacilitysaral2.org/mod/quiz/view.php?id=3458) |
| 3 | [Packages and the native pipe](primer-03-packages-pipe.qmd) | `library()`, and chaining steps with <code>&#124;&gt;</code> | [Primer 3](https://setfacilitysaral2.org/mod/quiz/view.php?id=3459) |
| 4 | [The five dplyr verbs](primer-04-dplyr-verbs.qmd) | `filter()`, `select()`, `arrange()`, `mutate()`, `summarise()`, and missing values | [Primer 4](https://setfacilitysaral2.org/mod/quiz/view.php?id=3460) |
| 5 | [Your first plot](primer-05-first-plot.qmd) | `ggplot()`: data, aesthetics, geom | [Primer 5](https://setfacilitysaral2.org/mod/quiz/view.php?id=3461) |

**Do those first.** Allow about 45 minutes for all five.

This page is the reference card for afterwards. It holds the four things the
primers do not: the type table, how to read a file from disk, the errors you
will meet, and where to read more.

::: {.callout-tip}
## Then do it again in RStudio
The browser primers cannot teach you file paths or projects, because there are
no files in a browser. Once R and RStudio are installed, retype a few of the
primer cells in your own console. Typing is slower than copy-paste, and that is
the point: you notice the brackets.
:::

## The four types you will meet

A vector, and therefore a tibble column, holds only **one** type.

| Type | Example | In your data |
|---|---|---|
| `numeric` | `72.5` | Marks, age, BMI |
| `character` | `"MBBS"` | Names, comments |
| `logical` | `TRUE` | Pass/fail flags |
| `factor` | `Female` | Categories with fixed levels |

Mixing types does not raise an error. R silently converts everything to the most
permissive type, which is a classic source of confusion:

```{r}
c(1, 2, "three")     # all become character
```

`str()` is the function to reach for when something goes wrong. Nine times out
of ten the problem is a column that is `character` when you thought it was
`numeric`.

::: {.callout-note}
## Tibble or data frame?
The primers say **tibble**; older books and most online answers say **data
frame**. A tibble *is* a data frame, with better printing and fewer surprises.
`data.frame()` is base R; `tibble()` comes from the tidyverse. Everything in
this workshop works on both.
:::

## Getting data in

```{r}
#| eval: false
library(tidyverse)
library(here)

students <- read_csv(here("data", "meded_students.csv"))
```

A relative path, because you are working inside an RStudio project. `here()`
builds the path from the project folder, so the same line works on your laptop
and on your colleague's. See [Datasets](data.qmd) for the files.

::: {.callout-warning}
## Never `setwd()`
`setwd("C:/Users/arun/Desktop/analysis")` is the one line guaranteed to break
your script on any other computer. An RStudio project plus `here()` replaces it.
:::

## Errors you will meet, and what they mean

R is case sensitive. Most beginner errors are a capital letter or a bracket.

| Message | Cause | Fix |
|---|---|---|
| `could not find function "read_csv"` | Package not loaded | `library(readr)`, or `library(tidyverse)` which includes it |
| `there is no package called 'readr'` | Package not installed | `install.packages("readr")`, in the console, once |
| `object 'viva' not found` | Typo, or you never ran the line that created it | Re-run the earlier line; check spelling and case |
| `unexpected symbol` / `unexpected ')'` | Missing comma or unbalanced bracket | Look at the line *above* the one flagged |
| `+` in the console instead of `>` | R is waiting for you to finish an expression | Press <kbd>Esc</kbd> and retype |
| A result of `NA` | A missing value upstream | Count them with `sum(is.na(x))`, then decide: `na.rm = TRUE` |

Read the message. It names which of the two things is missing, the function or
the object, and that tells you whether to load a package or fix a typo.

## Handy shortcuts

| Shortcut | Does |
|---|---|
| <kbd>Alt</kbd> + <kbd>-</kbd> | Types `<-` |
| <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> | Types the pipe. It gives <code>&#124;&gt;</code> only if you ticked **Use native pipe operator** during [setup](install.qmd) |
| <kbd>Ctrl</kbd> + <kbd>Enter</kbd> | Runs the current line |
| <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>F10</kbd> | Restarts R (a clean session) |
| <kbd>Tab</kbd> | Completes a name, and shows the arguments |

## Where to go if you want more

- [R for Data Science (2e), chapters 2–4](https://r4ds.hadley.nz/): free online, the standard reference
- [Posit primers](https://posit.cloud/learn/primers): interactive, browser-based, no installation
- [Swirl](https://swirlstats.com/): learn R inside R, `install.packages("swirl"); library(swirl); swirl()`

## Next

→ [Pre-reading](pre-reading.qmd)
