---
title: "R primer"
subtitle: "The 45 minutes of R that Day 1 assumes"
---

Day 1 starts with data *visualisation* at 11:15, which means you will be
handed a data frame and asked to plot it. This page makes sure the words
"data frame", "vector" and "package" mean something to you by then.

Work through it in RStudio. Type the code. Do not copy-paste. Typing is
slower, and that is the point: you notice the brackets.

## Objects: everything has a name

R stores things in objects. `<-` is the assignment arrow.

```{r}
n_students <- 240
course     <- "Community Medicine"
passed     <- TRUE

n_students
course
```

::: {.callout-tip}
## Shortcut
<kbd>Alt</kbd> + <kbd>-</kbd> types `<-` for you. Use it from the start;
your wrists will thank you.
:::

## Functions: verbs that take arguments

```{r}
sqrt(64)
round(3.14159, digits = 2)
mean(c(72, 68, 81, 90))
```

`c()` is the most-used function in R: it **c**ombines values into a vector.
Arguments in brackets, separated by commas. Named arguments (`digits = 2`)
are clearer than positional ones. Prefer them.

If you do not know what a function does, ask R:

```{r}
#| eval: false
?round
```

## Vectors: a column of values, all the same type

```{r}
scores <- c(72, 68, 81, 90, 55)
scores

length(scores)
mean(scores)
scores > 70          # a comparison returns a logical vector
```

The four types you will meet:

| Type | Example | In your data |
|---|---|---|
| `numeric` | `72.5` | Marks, age, BMI |
| `character` | `"MBBS"` | Names, comments |
| `logical` | `TRUE` | Pass/fail flags |
| `factor` | `Female` | Categories with fixed levels |

A vector holds **one** type. Mixing them silently converts everything to the
most permissive type, a classic source of confusion:

```{r}
c(1, 2, "three")     # all become character
```

## Data frames: the rectangle you actually analyse

A data frame is a set of equal-length vectors side by side: rows are
observations (students, patients, sessions), columns are variables.

```{r}
students <- data.frame(
  id     = c("S01", "S02", "S03", "S04"),
  batch  = c("2023", "2023", "2024", "2024"),
  score  = c(72, 68, 81, 90),
  passed = c(TRUE, TRUE, TRUE, TRUE)
)

students
```

Ways to look at one:

```{r}
nrow(students)     # rows
ncol(students)     # columns
names(students)    # variable names
str(students)      # structure: types and first values
summary(students)  # quick descriptives
```

`str()` is the function you will use most often when something goes wrong.
Nine times out of ten the problem is that a column is a `character` when you
thought it was a `numeric`.

Pull out a single column with `$`:

```{r}
students$score
mean(students$score)
```

## Packages: R's extensions

Base R can do a lot. The rest arrives in packages. Install once, load every
session.

```{r}
#| eval: false
install.packages("dplyr")   # once, ever
library(dplyr)              # every session, at the top of every script
```

::: {.callout-warning}
## `install.packages()` does not belong in your script
Put `library()` calls at the top of a script; keep `install.packages()` in
the console. A script that reinstalls packages every run is slow and rude to
whoever else runs it.
:::

## The pipe `|>`

The pipe passes the thing on its left into the function on its right as the
first argument. These two lines do exactly the same thing:

```{r}
round(mean(c(72, 68, 81, 90)), 1)

c(72, 68, 81, 90) |> mean() |> round(1)
```

The second reads left to right, in the order the work happens: *take these
scores, take the mean, round to one decimal*. Once your analysis is six steps
long, the pipe is the difference between readable and unreadable code. We use
`|>` (native, built into R 4.1+) throughout this workshop.

::: {.callout-note}
## `|>` vs `%>%`
You will see `%>%` (from the **magrittr**/**dplyr** world) in older code and
in most tutorials online. For everything in this workshop they behave the
same. We use `|>` because it needs no package.
:::

## Getting data in

```{r}
#| eval: false
library(readr)
students <- read_csv("data/meded_students.csv")
```

Relative path, because you are working inside an RStudio project. See
[Datasets](data.qmd) for the files.

## Exercise

::: {.callout-note title="Exercise: your first data frame"}
**Time:** 10 minutes

1. Create a vector `osce` with the marks 78, 64, 91, 55, 83.
2. Find its mean and its highest value (`max()`).
3. How many of those marks are 70 or above? *Hint:* `sum(osce >= 70)`.
   `TRUE` counts as 1.
4. Build a data frame `viva` with two columns: `station`
   (`"History"`, `"Exam"`, `"Counselling"`) and `marks` (`8`, `7`, `9`).
5. Print the structure of `viva`.
:::

::: {.callout-tip title="Solution" collapse="true"}
```{r}
osce <- c(78, 64, 91, 55, 83)

mean(osce)
max(osce)
sum(osce >= 70)

viva <- data.frame(
  station = c("History", "Exam", "Counselling"),
  marks   = c(8, 7, 9)
)

str(viva)
```

Note that `station` is a character column and `marks` is numeric, exactly
what `str()` tells you.
:::

## Errors you will meet, and what they mean

| Message | Cause | Fix |
|---|---|---|
| `could not find function "read_csv"` | Package not loaded | `library(readr)` |
| `object 'students' not found` | Typo, or you never ran the line that created it | Re-run the earlier line; check spelling and case |
| `unexpected symbol` / `unexpected ')'` | Missing comma or unbalanced bracket | Look at the line *above* the one flagged |
| `+` in the console instead of `>` | R is waiting for you to finish an expression | Press <kbd>Esc</kbd> and retype |

R is case sensitive. `Students` and `students` are different objects. So are
`TRUE` and `True`. The second is an error.

## Where to go if you want more

- [R for Data Science (2e), chapters 2–4](https://r4ds.hadley.nz/): free online, the standard reference
- [Posit primers](https://posit.cloud/learn/primers): interactive, browser-based, no installation
- [Swirl](https://swirlstats.com/): learn R inside R, `install.packages("swirl"); library(swirl); swirl()`

## Next

→ [Datasets](data.qmd): get the files we will use on both days.
