---
title: "7 · Publication-ready tables & figures"
subtitle: "gtsummary"
---

::: {.session-meta}
::: {.meta-item}
[Teaching]{.meta-label}[11:15 – 12:00]{.meta-value}
:::
::: {.meta-item}
[Hands-on]{.meta-label}[12:00 – 13:00]{.meta-value}
:::
::: {.meta-item}
[Faculty]{.meta-label}[Arun Mitra]{.meta-value}
:::
::: {.meta-item}
[Slides]{.meta-label}[[Open deck ↗](../slides/day2/s7-gtsummary.html){target="_blank"}]{.meta-value}
:::
:::

```{r}
#| include: false
library(tidyverse)
library(gtsummary)
theme_gtsummary_compact()
students <- read_csv("data/meded_students.csv", show_col_types = FALSE)
```

## Learning objectives

1. Produce a Table 1 with `tbl_summary()` in one line.
2. Stratify by a grouping variable and add an appropriate comparison test.
3. Control which statistics, labels and digits appear.
4. Tabulate a regression model with `tbl_regression()`.
5. Export a table to Word without reformatting it by hand.

## Table 1, in one line

```{r}
students |>
  select(age, sex, rural_background, attendance_pct, pretest, burnout_band) |>
  tbl_summary()
```

`gtsummary` has already decided, correctly, that `age` and `attendance_pct`
are continuous (median and IQR), and that `sex` and `burnout_band` are
categorical (n and %). It has counted the missing values and labelled them.

This is the table you have built by hand, in Word, for every paper you have
written.

## Stratified, with a test

```{r}
students |>
  select(teaching_arm, age, sex, rural_background, attendance_pct,
         pretest, burnout_band) |>
  tbl_summary(
    by = teaching_arm,
    missing_text = "Missing"
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels()
```

`add_p()` picks the test by variable type: Wilcoxon rank-sum for continuous,
chi-squared for categorical, Fisher's exact when expected counts are small.

::: {.callout-warning}
## `add_p()` is convenient, and convenience is a risk
It will happily produce a p-value for every row of your Table 1. Testing
baseline differences in a randomised comparison is
[widely criticised](https://doi.org/10.1002/sim.4780090113): if allocation was
random, any imbalance *is* chance by definition, and the test answers a
question nobody asked.

Report baseline characteristics with descriptive statistics and a comment on
clinically meaningful imbalance. Keep `add_p()` for the comparisons you
actually pre-specified.
:::

## Taking control

```{r}
students |>
  select(teaching_arm, age, attendance_pct, pretest, posttest, gain, sex) |>
  tbl_summary(
    by = teaching_arm,
    label = list(
      age            ~ "Age (years)",
      attendance_pct ~ "Attendance (%)",
      pretest        ~ "Pre-test score",
      posttest       ~ "Post-test score",
      gain           ~ "Knowledge gain",
      sex            ~ "Sex"
    ),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    missing = "no"
  ) |>
  add_p(test = all_continuous() ~ "t.test") |>
  modify_header(label ~ "**Characteristic**") |>
  modify_caption("**Table 1.** Cohort characteristics by teaching arm.") |>
  bold_labels()
```

The pieces:

| Argument | Controls |
|---|---|
| `label =` | Human-readable variable names. Do this always. `attendance_pct` is not acceptable in a manuscript |
| `statistic =` | `{mean} ({sd})` vs the default `{median} ({p25}, {p75})` |
| `digits =` | Decimal places |
| `missing =` | `"ifany"` (default), `"no"`, `"always"` |
| `add_p(test =)` | Override the automatic test choice |

::: {.callout-tip}
## Report the statistic your data deserve
The default is median and IQR because it is the safer choice for skewed
data. Switching to mean and SD is a claim about your distribution, so check
it with a histogram before you make it. `study_hours` in this dataset is
right-skewed; `pretest` is not.
:::

## Regression tables

```{r}
model <- lm(gain ~ teaching_arm + pretest + attendance_pct + sex,
            data = students)

tbl_regression(
  model,
  label = list(
    teaching_arm   ~ "Teaching arm",
    pretest        ~ "Pre-test score",
    attendance_pct ~ "Attendance (%)",
    sex            ~ "Sex"
  )
) |>
  add_glance_table(include = c(nobs, r.squared, adj.r.squared)) |>
  bold_p(t = 0.05) |>
  modify_caption("**Table 2.** Linear regression of knowledge gain.")
```

Coefficients, confidence intervals and p-values, formatted, labelled and
ready. For a logistic model, `tbl_regression(..., exponentiate = TRUE)` gives
odds ratios instead.

```{r}
glm(factor(passed) ~ teaching_arm + attendance_pct,
    data = students, family = binomial) |>
  tbl_regression(exponentiate = TRUE) |>
  bold_p()
```

## Side by side

```{r}
t1 <- students |> select(gain, teaching_arm) |>
  tbl_summary(by = teaching_arm, missing = "no")

t2 <- lm(gain ~ teaching_arm, data = students) |>
  tbl_regression()

tbl_merge(list(t1, t2),
          tab_spanner = c("**Descriptive**", "**Unadjusted model**"))
```

`tbl_stack()` does the same vertically, useful for reporting the same model
across several outcomes.

## Getting it into Word

```{r}
#| eval: false
tbl <- students |>
  select(teaching_arm, age, sex, pretest) |>
  tbl_summary(by = teaching_arm) |>
  add_p()

# Option 1, inside a Quarto document rendering to docx: just print it.

# Option 2: a standalone Word file
tbl |>
  as_flex_table() |>
  flextable::save_as_docx(path = "output/table1.docx")
```

::: {.callout-important}
## Option 1 is the point of the whole workshop
If your table lives in a chunk in your manuscript `.qmd`, then correcting a
data entry error means re-rendering, not rebuilding the table. Exporting to a
separate Word file re-introduces the manual step you spent two days learning
to remove. Use it only when a co-author insists on a bare table.
:::

# Hands-on {#hands-on}

**12:00 – 13:00.** Exercises 1–3; 4 is a stretch.

::: {.callout-note title="Exercise 1: your Table 1"}
**Time:** 15 minutes

1. Build a `tbl_summary()` of the cohort stratified by `college`.
2. Include `age`, `sex`, `rural_background`, `attendance_pct`, `pretest`.
3. Give every variable a proper label.
4. Add an overall column.
:::

::: {.callout-tip title="Solution 1" collapse="true"}
```{r}
students |>
  select(college, age, sex, rural_background, attendance_pct, pretest) |>
  tbl_summary(
    by = college,
    label = list(
      age              ~ "Age (years)",
      sex              ~ "Sex",
      rural_background ~ "Domicile",
      attendance_pct   ~ "Attendance (%)",
      pretest          ~ "Pre-test score"
    )
  ) |>
  add_overall() |>
  bold_labels()
```
:::

::: {.callout-note title="Exercise 2: outcomes with a test"}
**Time:** 15 minutes

1. Table of `pretest`, `posttest` and `gain` by `teaching_arm`.
2. Report mean (SD).
3. Add a t-test.
4. Add a caption.
5. What does the `pretest` row tell you, and why does it matter for how you
   read the `gain` row?
:::

::: {.callout-tip title="Solution 2" collapse="true"}
```{r}
students |>
  select(teaching_arm, pretest, posttest, gain) |>
  tbl_summary(
    by = teaching_arm,
    statistic = all_continuous() ~ "{mean} ({sd})",
    digits = all_continuous() ~ 1,
    label = list(pretest ~ "Pre-test", posttest ~ "Post-test",
                 gain ~ "Gain (post − pre)")
  ) |>
  add_p(test = all_continuous() ~ "t.test") |>
  modify_caption("**Table.** Knowledge scores by teaching arm.")
```

The `pretest` row should show no meaningful difference. The arms were
allocated alternately, so they start comparable. That is what licenses reading
the `gain` difference as an effect of teaching method rather than of who
ended up in which arm. Had the pre-test differed substantially, the gain
comparison would need adjustment, and you would have discovered that *before*
interpreting it.
:::

::: {.callout-note title="Exercise 3: a model"}
**Time:** 15 minutes

1. Fit `lm(posttest ~ teaching_arm + pretest + attendance_pct)`.
2. Tabulate it with `tbl_regression()`, labelled.
3. Add model fit statistics.
4. Interpret the `teaching_arm` coefficient in one sentence a clinician would
   understand.
:::

::: {.callout-tip title="Solution 3" collapse="true"}
```{r}
lm(posttest ~ teaching_arm + pretest + attendance_pct, data = students) |>
  tbl_regression(
    label = list(teaching_arm   ~ "Teaching arm",
                 pretest        ~ "Pre-test score",
                 attendance_pct ~ "Attendance (%)")
  ) |>
  add_glance_table(include = c(nobs, r.squared)) |>
  bold_p()
```

Interpretation: *adjusting for pre-test score and attendance, students taught
by flipped classroom scored about 5 points higher on the post-test than those
taught by traditional lecture.* Note the two things that sentence contains
which a bare p-value does not: the size of the effect, and what it is
adjusted for.
:::

::: {.callout-note title="Exercise 4 (stretch): merge and export"}
**Time:** 15 minutes

1. Build a descriptive table of `gain` by arm and a regression table of
   `gain ~ teaching_arm`.
2. `tbl_merge()` them with sensible spanners.
3. Export the result to `output/table_combined.docx`.
4. Open it in Word. What still needs manual work?
:::

::: {.callout-tip title="Solution 4" collapse="true"}
```{r}
#| eval: false
t1 <- students |> select(gain, teaching_arm) |>
  tbl_summary(by = teaching_arm, statistic = all_continuous() ~ "{mean} ({sd})",
              missing = "no")
t2 <- lm(gain ~ teaching_arm, data = students) |> tbl_regression()

tbl_merge(list(t1, t2),
          tab_spanner = c("**Descriptive**", "**Unadjusted**")) |>
  as_flex_table() |>
  flextable::save_as_docx(path = "output/table_combined.docx")
```

What still needs manual work: journal-specific fonts and table borders, and
sometimes the footnote wording. Set those once in a `reference-doc` Word
template rather than fixing them after every render, see
[Session 6](s6-quarto-communication.qmd).
:::

## Key takeaways

::: {.callout-note appearance="minimal"}
- `tbl_summary()` builds Table 1; `by =` stratifies it; `add_p()` tests it.
- Label every variable. `attendance_pct` belongs in your code, not your
  manuscript.
- Default is median (IQR); mean (SD) is a claim about your distribution.
- Do not p-test baseline characteristics of a randomised comparison.
- `tbl_regression()` for models; `exponentiate = TRUE` for odds ratios.
- Keep tables inside the `.qmd`. Exporting to Word re-introduces the manual
  step.
:::

## Next

→ [8 · Generative AI in health research](s8-genai.qmd)
