---
title: "Publication-ready tables & figures"
subtitle: "gtsummary · Day 2, Session 7"
author: "Arun Mitra"
date: 2026-09-10
format:
  revealjs:
    theme: [default, ../custom.scss]
    slide-number: c/t
    footer: "MedEd Conclave 2026 · R & Reproducible Research"
    transition: fade
    center: false
execute:
  echo: true
  warning: false
---

```{r}
#| include: false
library(tidyverse); library(gtsummary)
theme_gtsummary_compact()
students <- read_csv("data/meded_students.csv", show_col_types = FALSE)
```

## Table 1, in one line {.section-divider}

```{r}
#| output-location: column
students |>
  select(age, sex, rural_background,
         pretest, burnout_band) |>
  tbl_summary()
```

::: {.notes}
⏱️ 3 min. Let it land in silence. Every person in the room has built this
table by hand, more than once.
:::

## Stratified, tested

```{r}
#| output-location: column
students |>
  select(teaching_arm, age, sex,
         attendance_pct, pretest) |>
  tbl_summary(by = teaching_arm) |>
  add_p() |>
  add_overall() |>
  bold_labels()
```

## But: don't p-test baseline

If allocation was random, imbalance **is** chance by definition.

. . .

The test answers a question nobody asked.

Describe the baseline. Test the outcome you pre-specified.

::: {.notes}
🎯 Altman 1985. Expect pushback — many local journals still demand it. Advice:
give them the column, and say in the text why it is uninformative.
:::

## Take control

```r
tbl_summary(
  by = teaching_arm,
  label = list(attendance_pct ~ "Attendance (%)"),
  statistic = all_continuous() ~ "{mean} ({sd})",
  digits = all_continuous() ~ 1
)
```

`attendance_pct` belongs in your code. Not in your manuscript.

## Models

```{r}
#| output-location: column
lm(gain ~ teaching_arm + pretest,
   data = students) |>
  tbl_regression() |>
  add_glance_table(
    include = c(nobs, r.squared)) |>
  bold_p()
```

`exponentiate = TRUE` → odds ratios.

## Into Word

**Option 1** — table lives in the manuscript `.qmd`. Re-render.

**Option 2** — `as_flex_table() |> save_as_docx()`.

. . .

Option 2 re-introduces the manual step you spent two days removing. Use it
only when a co-author insists.

## Hands-on — 60 min {background-color="#0f5257"}

Your Table 1, your outcome table, your model.

::: {.notes}
📋 Common: forgetting `by =` is a column not a filter; label lists needing
`~` not `=`.
:::
